---
title: "Exploratory Data Analysis in R"
author: "Ivailo Kassamakov"
date: "February 2, 2016"
output: html_document
---

# Abstract
This project is an effort towards partial fullfilment of the requirements of Udacity's Data Analyst nanodegree.

The purpose is to perform an Exploratory Data Analysis (mono-, bi- and multivariate) on a dataset containing chemical measurements and degustation results of a sample of red wines.

# Goal of EDA
Our goal with this dataset is quite naturally trying to discover a model that could allow us to predict the quality of the wine given its chemical characteristics.

Of course, in the meantime, we will discover any interaction between the "chemical variables".

# Structure of the dataset

We begin by loading all needed packages:
```{r Package loading}
suppressMessages(library(ggplot2))
suppressMessages(library(gridExtra))
suppressMessages(library(dplyr))
suppressMessages(library(GGally))
```

Next, we load the dataset (stored in a CSV file) into a data.frame:

```{r Loading of the dataset}
w <- read.csv("wineQualityReds.csv")
```

Some information about the number and the structure of the observations
```{r Dataset structure}
str(w)
```

We see that there are 1599 observation of wines, each one containing 13 variables. Currently, all variables are numeric. This is valid for most of them, since they are results of laboratory measurements. However, the "quality" variable is the degustation result, which logically forms an ordinal scale. Therefore, we can convert this variable to an ordered factor, assuming that a higher number indicates a higher quality

```{r Converting "quality" to an ordered factor}
w$quality <- factor(w$quality, ordered = T)
str(w)
```

We clearly see that "quality" is now represented by a 6-level ordered factor. 

We can also drop the column named "X", as it simply repeats the numeric index of the observations:

```{r Removing "X" column}
w$X <- NULL
```

# Overview statistics of the dataset

Next we obtain the standard 6-number statistics for all numeric variables, together with the distribution of the "quality" factor.

```{r Statistical summary}
summary(w)
```




